---
title: "Northwest Conference Women's Soccer Rankings"
date: "2/1/2023"
output: 
  html_document:
    toc: true
    toc_float: true
params:
  data_dir: "_data"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, error = FALSE)
library(tidyverse)
library(here)
library(directlabels) # per suggestion in this answer: https://stackoverflow.com/a/29359816
library(plotly)
```

## Background

When I played for the Whitman College women's soccer team from 2010-2013, Puget Sound dominated the [Northwest Conference](https://nwcsports.com/index.aspx). However, when I checked the rankings today, Pacific Lutheran had beat out Puget Sound for the top position. I was curious to see how team standings had changed year-over-year. 

```{r data_read_standings}
file <- tail(list.files(here(params$data_dir), pattern = "nwc_rankings_wsoc", full.names = TRUE), 1)
standings <- read_csv(file, col_types = cols())
```

```{r data_clean_standings}
standings_clean <- standings %>%
  # put record into separate columns
  separate(conf, sep = "-", into = c("wins", "losses", "draws"), remove = FALSE, fill = "right", convert = TRUE) %>%
  mutate(draws = ifelse(is.na(draws), 0, draws)) %>%
  # calculate points
  rowwise() %>%
  mutate(
    pts_calculated = sum(wins * 3 + draws)
  ) %>%
  ungroup() %>%
  # add ranking
  group_by(standings_year) %>%
  mutate(rank = rank(-pts_calculated)) %>%
  ungroup()

# ensure calculated points matches reported points
stopifnot({
  standings_clean %>%
    filter(!is.na(pts)) %>%
    filter(pts != pts_calculated) %>%
    {nrow(.) == 0}
})
```

## Exploration

During my career (2010-2013), Puget Sound won the league handily. However, while Puget Sound has continued to perform at the top of the league since then, Pacific Lutheran has had a dramatic improvement in performance and has secured first place since 2016*. 

*Due to COVID, the 2020 season was [postponed](https://www.d3sports.com/notables/2020/06/schools-call-it-off-for-fall). Although it looks like some games may have been played in the spring of 2021, the standings weren't reported on the official Northwest Conference site, as far as I can tell. 

```{r plot_pts_standings}
YEARS <- seq.int(
  min(standings_clean$standings_year), 
  max(standings_clean$standings_year), 
  length.out = 5)
hallie_years <- data.frame(
  start_year = 2010,
  end_year = 2013,
  annotation = "Hallie's Career"
)
plot_pts_standings <- standings_clean %>%
  ggplot(aes(x = standings_year, y = pts_calculated, color = school)) + 
  theme_bw() + 
  coord_cartesian(ylim = c(0, max(standings_clean$pts_calculated))) + 
  scale_x_continuous(breaks = YEARS) + 
  theme(panel.grid.minor.x = element_blank()) + 
  geom_rect(
    data = hallie_years,
    inherit.aes = FALSE,
    aes(
      color = annotation,
      xmin = start_year, xmax = end_year),
    ymin = -5, ymax = max(standings_clean$pts_calculated)*1.1,
    alpha = 0.25
  ) +
  geom_point() +
  geom_line()
ggplotly(plot_pts_standings)
```

*Click on a team in the legend to show/hide its data. Double click on a team to only show its data.*

```{r data_read_coaches}
# manually create ups coaches
coaches_ups <- bind_rows(
  data.frame(year = 1995:2015, head_coach = "Randy Hanson"),
  data.frame(year = 2016:2018, head_coach = "Joseph Vari"),
  data.frame(year = 2019:2022, head_coach = "Kim Calkins")
) %>%
  mutate(school = "Puget Sound", 
         school_abbrev = "UPS")

# read in plu coaches and clean
filename <- list.files(here(params$data_dir), pattern = "plu_wsoc_head_coaches_", full.names = TRUE)
coaches_plu <- read_csv(filename, col_types = cols()) %>%
  select(year, head_coach = `head coach`) %>%
  # add 2022 season
  bind_rows(data.frame(year = "2022", head_coach = "Seth Spidahl")) %>%
  mutate(school = "Pacific Lutheran", 
         year = as.numeric(year), 
         school_abbrev = "PLU") %>%
  # removes 2020 season
  filter(!is.na(year))

# combine
coaches <- bind_rows(coaches_ups, coaches_plu)
```

Pacific Lutheran's meteoric rise in rankings coincides with the hiring of a new coach, Seth Spidahl. Puget Sound's change in rankings doesn't seem to coincide as cleanly with a single coach. 

```{r data_plot_coaches}
coaches_standings <- inner_join(
  standings_clean, coaches, by = c("standings_year" = "year", "school")
) %>%
  arrange(school, standings_year) %>%
  mutate(
    school_head_coach = sprintf("%s: %s", school_abbrev, head_coach),
    school_head_coach = factor(school_head_coach, levels = unique(school_head_coach))
  )
plot_coaches_standings <- coaches_standings %>%
  ggplot(aes(x = standings_year, y = pts_calculated)) +
  geom_line(aes(group = school, color = school_head_coach)) +
  geom_point(aes(color = school_head_coach), size = 4) +
  theme_bw() + 
  coord_cartesian(ylim = c(0, max(standings_clean$pts_calculated))) + 
  scale_color_manual(values = c("gold", "gold3", "gold4", "firebrick1", "firebrick3", "firebrick4")) +
  scale_x_continuous(breaks = YEARS) + 
  labs(title = "Northwest Conference Women's Soccer - Head Coaches")
print(plot_coaches_standings)
```

## Conclusion

Pacific Lutheran has placed first in the Northwest Conference Women's Soccer league for the last six years, displacing Puget Sound as the dominant team since I played. The hiring of a new coach at Pacific Lutheran coincides with this improvement in performance.

## Data Sources

### Conference Standings

The women's soccer standings are hosted on the [Northwest Conference site](https://nwcsports.com/standings.aspx?path=wsoc). I [scraped](https://github.com/hallieswan/hallieswan.github.io/blob/main/_data_prep/nwc_rankings.R) the standings for each year using [rvest](https://rvest.tidyverse.org).

### Head Coaches

The head coaches for Pacific Lutheran [up to 2021](https://golutes.com/sports/2018/5/30/womens-soccer-year-by-year-history.aspx) was [easily scraped](https://github.com/hallieswan/hallieswan.github.io/blob/main/_data_prep/nwc_wsoc_head_coaches_plu.R). I manually curated the head coach for [2022](https://golutes.com/sports/womens-soccer/coaches/2022).

Puget Sound was a bit more difficult, because I couldn't find a well structured data source. I manually curated a list of coaches --  

  - Randy Hanson was head coach from [1995-2015](https://trail.pugetsound.edu/?p=13207)
  - Joseph Vari became head coach in [2016](https://trail.pugetsound.edu/?p=14508) and continued through [2018](https://nwcsports.com/news/2018/11/7/11_7_2018_4606.aspx)
  - Kim Calkins has been head coach [since 2019](https://loggerathletics.com/sports/womens-soccer/roster/coaches/kim-calkins/160)
